---
title: "Data visualization assignment"
author: "Samantha Bunner"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE, fig.width = 6, fig.height = 4)
```

In this exercise, you will practice basic plotting with **`ggplot2`** on
a dataset of median gene expression levels across samples from various
human tissues from the GTEx Project. The dataset (`dt`) has one row per
gene, with columns:

- `Name`: the Ensembl gene ID
- `Description`: the gene symbol
- One column per tissue containing the median expression levels (in TPM
  units)

Insert your code into the empty code blocks. Insert answers to any
comprehension questions as comments within the code blocks. You do not
need to submit the figures themselves, only the code used to produce
them. Remember to label your axes!

**Using HopGPT.** As in the last assignment: you are welcome to use it,
and wherever you do, add a couple of comment lines in that same code
block noting what you asked it and how you checked the answer.

------------------------------------------------------------------------

### Load data

Start by loading the data into R, as before.

```{r load-data}
library(tidyverse)

dt <- read_delim("/Users/cmdb/Data/GTEx/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
                 skip = 2, col_names = TRUE, delim = "\t")
```

Recreate the `is_mitochondrial` column from the last assignment; you
will need it in Q4 and Q5.

```{r add-mito}
dt <- dt %>%
  mutate(is_mitochondrial = str_starts(Description, "MT-"))

dt %>% count(is_mitochondrial)
```

------------------------------------------------------------------------

### Q1. Histogram

Make a histogram of gene expression in the liver. Use the argument
`breaks = seq(0, 100, 2)` to bin in units of 2 TPM, ranging from 0 to
100. Note that this cuts off the data at 100, which might be misleading
depending on what we are trying to depict! How many genes are you
leaving out?

What do you notice about the shape of the distribution? What does it
tell you about the relative expression of genes in the liver
transcriptome? Do you think this is specific to the liver? Check another
tissue, as well.

```{r q1}

#nrow(dt)

ggplot(dt, aes(x = Liver)) +
  geom_histogram(breaks = seq(0, 100, 2)) +
  labs(
    x = "Expression (TPM)",
    y = "Counts",
    title = "Liver") +
  theme_minimal()

#I asked hop to explain the breaks part of the plot 

# 594 many genes are left out by making the range 0 to 100
Liver <- dt['Liver']
cat("Genes left out:", sum(Liver >100, na.rm = TRUE))

# answer here
# Most genes have very low levels of expression (0-2 TPM). The data is skewed to the right because the liver transcriptome probably needs the expression of a set of genes. I don't expect this to be specific to the liver, ill try the Bladder, which looks very similar.

ggplot(dt, aes(x = Bladder)) +
  geom_histogram(breaks = seq(0, 100, 2)) +
  labs(
    x = "Expression (TPM)",
    y = "Counts",
    title = "Bladder")+
  theme_minimal()

#
```

------------------------------------------------------------------------

### Q2. Density plot

Going back to the liver, now represent the same data with a density plot
rather than a histogram. Do not restrict the X axis. Within the same
code block, create a second plot where you remove the custom bin breaks,
but transform the expression values using `log10()`. First add a
pseudo-count of 1 to the expression values, as the log of 0 is negative
infinity.

What does the log-transformed version show you that the first plot does
not? And where on the log-transformed plot do the genes with 0 TPM end
up?

```{r q2}

# Not restricted x axis
ggplot(dt, aes(x = Liver)) +
  geom_density() +
  labs(
    x = "Expression (TPM)",
    y = "Density")

#Log-transformed
ggplot(dt, aes(x = log10(Liver + 1))) +
  geom_density() +
  labs(
    x = "log10(Expression + 1)",
    y = "Density"
  ) +
  theme_minimal()

# answer here
#before log10 transformation you cannot see the distribution very well because most counts are close to zero. Plotting the log10 makes the data easier to visualize, to see that the data is largely right skewed. The genes with 0 TPM stay 0 because log10(0+1) = 0
```

------------------------------------------------------------------------

### Q3. Scatter plot

Make a scatter plot comparing the expression of genes in the `Liver` to
the same genes in the `Heart - Left Ventricle`. Add a pseudocount to
each and log10 transform them. Adjust the point size to 1 and the alpha
(opacity) to 0.2. Add a dashed red line along the diagonal (y = x).

What does a point far above the diagonal mean, biologically?

```{r q3}

print(colnames(dt)[35]) #This is the Heart- Left Ventricle column

#I had trouble plotting by the name Heart - Left Ventricle so i used the column number instead. i made a few errors y axis outisde of aes, missing closing parenthesis that Hop helped with

ggplot(dt, aes(x = log10(Liver + 1), y = log10(.data[[names(dt)[35]]] + 1))) + 
  geom_point(alpha = 0.2, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Liver Expression",
    y = "Heart - Left Ventricle Expression"
  ) +
  theme_classic()
# answer here
# a point far above the diagonal means that gene is more highly expresseed in the Heart than in the Liver.
```

------------------------------------------------------------------------

### Q4. Colored points on scatter plot

Color the previous scatter plot points based on whether the genes are
mitochondrial vs. nuclear encoded, using the `is_mitochondrial` column
you made above.

There are only a few dozen mitochondrial genes among tens of thousands
of nuclear ones, so the colored points end up buried. Try plotting them
as a second `geom_point()` layer on top of the others, using the `data`
argument to give that layer just the mitochondrial genes:
`geom_point(data = filter(dt, is_mitochondrial), ...)`.

Where do the mitochondrial genes sit relative to the diagonal, and does
that fit what you found in the last assignment?

```{r q4}

#Color the previous scatter plot points based on whether the genes are
#mitochondrial vs. nuclear encoded, using the `is_mitochondrial` column you made above.


ggplot(dt, aes(
  x = log10(Liver + 1), 
  y = log10(.data[[names(dt)[35]]] + 1)
)) + 
  geom_point(color = "grey", alpha = 0.2, size = 1) +
  geom_point(data = dt %>% filter(is_mitochondrial), color = "purple", alpha = 0.2, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Liver Expression (log10(TPM + 1))",
    y = "Heart - Left Ventricle Expression (log10(TPM + 1))",
    title = "MT expression in Liver vs Heart"
  ) +
  theme_classic()





# answer here
#MT genes are above the Line which suggests expression is slightly higher in the heart than liver, which makes sense since heart muscle needs much more of the MT powerhouse of the cell than lazy liver
```

------------------------------------------------------------------------

### Q5. Facets

Repeat question 1 (histogram), but faceting such that you create
separate panels (arranged horizontally) for several tissues of interest:

```{r focal-tissues}
focal_tissues <- c("Liver", "Testis", "Whole Blood", "Heart - Left Ventricle", "Brain - Cortex")
```

This requires first converting the data into "long" format with
`pivot_longer()`:

```{r pivot-long}
dt_long <- dt %>%
  select("Description", "is_mitochondrial", all_of(focal_tissues)) %>%
  pivot_longer(cols = all_of(focal_tissues), names_to = "Tissue", values_to = "TPM")
```
After creating the 5-panel plot, also split into two rows that separate
genes that are versus are not encoded by the mitochondrial genome (10
panels). Use the argument `scales = "free_y"` to deal with the fact that
there are far fewer mitochondrial-encoded than nuclear-encoded genes.

`scales = "free_y"` means the panels can no longer be compared by eye.
Write the one-sentence note you would put in a figure caption to keep a
reader from being misled.

```{r q5}

# your code for 5-panel plot

dt_long %>%
  filter(Tissue %in% focal_tissues) %>% #Looks for rows in Tissues column if it matches the ones in the vector focal_tissues
  ggplot(aes(x = TPM)) +
  geom_histogram(breaks = seq(0, 100, 2)) +
  facet_wrap(~ Tissue) + 
  labs(
    x = "Expression (TPM)",
    y = "Counts"
  ) +
  theme_classic()


# your code for 10-panel plot 

dt_long %>%
  filter(Tissue %in% focal_tissues) %>%
  ggplot(aes(x = TPM)) +
  geom_histogram(breaks = seq(0, 100, 2)) +
  facet_grid(is_mitochondrial~ Tissue,scales = "free_y") + #For every combination
  labs(
    x = "Expression (TPM)",
    y = "Counts"
  ) +
  theme_classic()

# caption note
#Due to less MT genes, please note that the y axis are adjusted to visualize trends of each group (TRUE and FALSE)

```

------------------------------------------------------------------------

Optional: Have fun and explore these data! Create your own plots and
interpret them.

```{r optional}

# answer here
# no thank you!
```
